---
title: "Northwestern University Methods Workshop 2024"
author: "Basak TARAKTAS"
date: "2024-03-04"
output:
  html_document: default
  pdf_document: default
---

# Defining a network with igraph

Install igraph and igraphdata

```{r}
#install.packages('igraph')
#install.packages('igraphdata')

library(igraph)
library(igraphdata)
```

## 

### Download files from Github using readr

```{r}

## Set working directory: session > set working directory > choose a folder

library (readr)
vertices<-"https://raw.githubusercontent.com/enaline/northwestern-network-analysis-workshop/main/vertex_list.csv"
vertices <- read_csv(url(vertices), col_names = TRUE)
vertices


edgelist<- "https://raw.githubusercontent.com/enaline/northwestern-network-analysis-workshop/main/edgelist.csv"
data <- read_csv(url(edgelist), col_names = TRUE)
data


friendship<-"https://raw.githubusercontent.com/enaline/northwestern-network-analysis-workshop/main/fr.csv"
fr <- read_csv(url(friendship), col_names = FALSE)
fr

```

### Read in files after manual download

```{r}

vertices <- read.csv("vertex_list.csv", header = TRUE)
data <- read.csv("edgelist.csv", header = TRUE)
fr <- read.csv("fr.csv", header=T, row.names=1)



```

## Define a network from an adjacency matrix

```{r}

####Let's save the fr file as a matrix to build the friendship network
fr<- as.matrix(fr)

net <- graph_from_adjacency_matrix(fr)

plot(net)
```

Let us visualize

```{r}
##change vertex color using vertex.color =''
plot(net, vertex.color = "purple")
plot(net, vertex.color = "red")
plot(net, vertex.color = "blue")
```

## Try with different layouts

```{r}

##try with different layouts to improve visualization
plot(net, vertex.size=20, layout=layout.fruchterman.reingold(net) )
plot(net, vertex.size=20, layout=layout_as_star(net))
plot(net, vertex.size=20, layout=layout_as_tree(net))
plot(net, vertex.size=20, layout=layout.circle(net))
plot(net, vertex.size=20, layout=layout.grid(net))
plot(net, vertex.size=20, layout= layout.kamada.kawai(net))

```

## Define network from an edgelist and vertex list

Great for large datasets

```{r}

##save the edge list as a matrix (do not forget)
data<- as.matrix(data)


###construct the network
friends <- graph_from_edgelist(data[, 1:2], directed= F)

##Construct a directed network this time
#friends_dir <- graph_from_edgelist(data[, 1:2], directed= T)

plot(friends)

```

Let us set attributes and visualize

Vertices may have different attributes like observations in a dataset have (e.g., partisanship, gender for respondents in survey data)

```{r}
###convert tibble into data frame to refer to it when defining attributes
vertices <- as.data.frame(vertices)

###Define edge attributes: frequency of contact
E(friends)$frequency <- as.integer(data[,3])


###define vertex attributes: gender
V(friends)$gender <- as.factor(vertices[,2])


###You can visualize the network to highlight certain characteristics of interest in those attributes
plot(friends, 
     edge.color = ifelse(E(friends)$frequency >= 6, "purple", "green"), ##if contact freq is greater or equal than 6 set edge color to purple and green otherwise
     vertex.color = ifelse(V(friends)$gender =="nonbinary", "blue", "orange"), ###if the node is of gender male, set vertex color to blue, pink otherwise
     vertex.shape= ifelse(V(friends)$gender =="male","circle", "square")) ###if the node is of gender male, set vertex shape to blue, pink otherwise


```

# Analyzing graph structure

Visualization is not helpful for large graphs. Graph- and node-level metrics alone help us understand the structure of both large and helpful graphs.

For any graph, we want to know

a)  *graph size* (number of edges) & *graph order* (number of nodes)\
b)  *edge distribution* across the graph to understand whether

-   the graph is highly/sparsely connected

-   edge accumulation predominantly occurs around specific nodes, or if they are relatively evenly distributed throughout the graph

c)  *important nodes*

## Case Study: Zachary's Karate Club

```{r}

###Zachary's karate club is a built-in dataset in the igraphdata package. Activate the package
library(igraphdata) 

###load data
data("karate")


##plot to get a sense (optional)
plot(karate)

```

### a) Basic statistics about graph structure

```{r}
###get the list of vertices
V(karate) 


##get the number of nodes
gorder(karate) ###also vcount()

###get the number of edges
gsize(karate)

```

### b.1) Is the Karate Club a highly connected graph?

```{r}

#density
edge_density(karate)
```

### b. 2) Are most edges accumulated around certain nodes? Are edges somewhat equally/uniformly distributed across the graph?

#### Degree

```{r}
##get degree of the vertices in a graph 
degree(karate)


##min, max, mean degree
max(degree(karate))
min(degree(karate))
mean(degree(karate))


##let us create a degree object (deg) to use for visualization purposes
deg<- degree(karate)

###if directed network, specify the mode for in-degrees and out-degrees.
###Karate is undirected. Let's look at the friendship network instead (see above)
#deg_fr<- degree(friends_dir, mode="in")
#deg_fr<- degree(friends_dir, mode="out")



##get the degree sequence in descending order
#sort(degree(karate)) ##### sort degrees of each vertex
#sort(degree(karate), decreasing = FALSE) #### sort in an ascending order

##find the node with max degree
#which.max(deg) 


##Create a histogram
## hist(deg) 
hist(deg, breaks=1:vcount(karate)-1, main="Histogram of node degree") ###a cleaner histogram
##breaks defines the number and width of the bins
##1:vcount(friends)-1 means that there will be as many bins as the total number of vertices minus one


```

Let's create a visual representation of the graph based on vertex degrees.

```{r}

plot(karate, vertex.size = degree(karate)*0.8)

# Set edge color to gray, and the node color to orange.
# additional settings e.g., edge.arrow.size=.2, edge.curved=0

plot(karate, 
     vertex.color="purple", 
     vertex.size=ifelse(degree(karate)>mean(degree(karate)), degree(karate)*0.8, degree(karate)*0.4 ), ###make high degree nodes bigger than others
     vertex.label.color="black", 
     vertex.label.cex=.7)


```

```{r}


```

### c) Centrality: which actors play an important role in this club?

#### c.1. Degree centrality: based on sheer number of connections

```{r}

##### degree centrality

c= centr_degree(karate)  ###prints 
c$res #node-level centrality scores.
c$centralization #The graph level centrality index. measure of how centralized the network is around some focal nodes. A centralization score of 1 indicates a graph is like a star network, whereas 0 indicates a wheel

```

#### c.2. Eigenvector centrality: important connections rather than the mere number of connections

```{r}
eigen_centrality(karate) ####eigenvector centrality
sort(eigen_centrality(karate)$vector)   #### $vector pulls eigen scores for vertices 



###which actor has the highest eigenvector centrality
V(karate)$eigen <- eigen_centrality(karate)$vector ###first define eigen. centrality as a vertex attribute
which(V(karate)$eigen ==max(V(karate)$eigen)) ##identify that node with the which function
###notice this returns the index

V(karate)[34] ##this prints out the name of the actor


```

#### c.3. Betweenness centrality: Brokers & bridges

```{r}
betweenness(karate)   #### gives betweenness centrality scores of nodes
max(betweenness(karate))


V(karate)$betweenness <- betweenness(karate)

which(V(karate)$betweenness ==max(betweenness(karate)) )

V(karate)[1]
```
